---
title: End-Semester Project
subtitle: Bivariate Analysis on GDP, Sanitation and Life Expectancy across Nations in 2010
author: 
  - Raj Pratap Singh [BS2219]
  - Shreyansh Mukhopadhyay [BS2147]
  - Aman Das [BS2206]
format:
  revealjs:
    theme: [default, custom.scss]
    transition: slide
    controls: true
    controls-layout: bottom-right
    controls-back-arrows: faded
    hide-inactive-cursor: true
code-fold: true
code-summary: Source
highlight-style: gruvbox-dark
menu: false
---

```{r include=FALSE}
library(knitr)
library(ggplot2)

```

# Introduction

## Overview

This presentation demonstrates the capabilities of **Bivariate Analysis** on datasets, to infer relationship between various features of the Population. The features of interest today are:

::: {.incremental}
- **Logarithm of GDP per Capita**: Logarithm of Gross Domestic Product divided by Mid-year. Population. [*lngdp*]
- **Basic Sanitation Access**: Percentage of people using at least basic Sanitation facilities. [*snt*]
- **Life Expectancy**: The average number of years a newly born Child would live. [*lfx*]
:::

## Data

```{r echo=TRUE, collapse=TRUE}
script.dir <- getSrcDirectory(function(x) {x})
setwd(script.dir)

numerise = function(x){
  x[grepl("k$", x)] <- as.numeric(sub("k$", "", x[grepl("k$", x)]))*10^3
  x <- as.numeric(x)
  return(x)
}

d1_raw = read.csv("./Data/gdp.csv")
d2_raw = read.csv("./Data/sanitation.csv")
d3_raw = read.csv("./Data/life_expectancy.csv")

yearname = "X2010"

d1 = d1_raw[!is.na(numerise(d1_raw[, yearname])),][,c("country", yearname)]
colnames(d1)[2] = "lngdp"
d2 = d2_raw[!is.na(numerise(d2_raw[, yearname])),][,c("country", yearname)]
colnames(d2)[2] = "snt"
d3 = d3_raw[!is.na(numerise(d3_raw[, yearname])),][,c("country", yearname)]
colnames(d3)[2] = "lfx"

dtemp = merge(x = d1, y = d2, by = "country")
d = merge(x = dtemp, y = d3, by = "country")

d$lngdp = log(numerise(d$lngdp))

write.csv(d, "./Data/assembled.csv")

kable(head(d))
```

# Bivariate Analysis

##

## Scatter Plot

A `scatter plot` (also called a `scatterplot`, `scatter graph`, `scatter chart`, `scattergram`, or `scatter diagram`) is a type of **Plot** or **mathematical diagram** using *Cartesian coordinate system* to display values for typically two Variable for a set of data. If the points are coded (color/shape/size), one additional variable can be displayed.

---

### Sanitation vs. GDP per Capita
``` {r echo=TRUE}
ggplot(d, aes(x = lngdp, y = snt))+
  geom_point()+
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "#f9f5d7"),
    panel.grid = element_line(colour = "#d5c4a1"),
    axis.line = element_line(colour = "#928374")
    )
```

---

### Life Expectancy vs. GDP per Capita
```{r echo=TRUE}
ggplot(d, aes(x = lngdp, y = lfx))+
  geom_point()+
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "#f9f5d7"),
    panel.grid = element_line(colour = "#d5c4a1"),
    axis.line = element_line(colour = "#928374")
    )
```

---

### Sanitation vs. Life Expectation
```{r echo=TRUE}
ggplot(d, aes(x = snt, y = lfx))+
  geom_point()+
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "#f9f5d7"),
    panel.grid = element_line(colour = "#d5c4a1"),
    axis.line = element_line(colour = "#928374")
    )
```