---
title: End-Semester Project
subtitle: Bivariate Analysis on GDP per capita, Sanitation and Life Expectancy across Nations in 2010
author:
  - Raj Pratap Singh [BS2219]
  - Shreyansh Mukhopadhyay [BS2147]
  - Aman Das [BS2206]
format:
  revealjs:
    width: 1200
    height: 750
    theme: [default, custom.scss]
    highlight-style: gruvbox
    transition: slide
    # logo: ./Data/logo.png
    hide-inactive-cursor: true
    code-fold: true
    menu: false
    scrollable: true
    history: true
  html:
    toc: true
    html-math-method: katex
    theme:
      light:
        - sandstone
      dark:
        - slate
    highlight-style: gruvbox
execute:
  cache: false
---

```{r include=FALSE}
library(knitr)
library(ggplot2)
library(quantreg)
```

# Introduction

## Overview

This presentation demonstrates the capabilities of *Bivariate Analysis* on datasets, to infer relationship between various features of Nations.

- **log of GDP per capita**: Logarithm (base $e$) of Gross Domestic Product (in $) per citizen. Adjusted for Inflation. [*lngdp*]
- **Sanitation Access %**: Percentage of people using at least basic Sanitation facilities, not shared with other households. [*snt*]
- **Life Expectancy**: The average number of years a newly born child would live, provided current mortality patterns hold. [*lfx*]

## Data

```{r echo=TRUE, collapse=TRUE}
script.dir = "~/Documents/Indian_Statistical_Institute/Sem_1/Statistical_Methods_1/Projects/Final_Sem/Formal"
setwd(script.dir)

numerise = function(x){
  x[grepl("k$", x)] <- as.numeric(sub("k$", "", x[grepl("k$", x)]))*10^3
  x <- as.numeric(x)
  return(x)
}

d1_raw = read.csv(file.path(".","Data","gdp.csv"), fileEncoding = 'UTF-8-BOM')
d2_raw = read.csv(file.path(".","Data","sanitation.csv"), fileEncoding = 'UTF-8-BOM')
d3_raw = read.csv(file.path(".","Data","life_expectancy.csv"), fileEncoding = 'UTF-8-BOM')

yearname = "X2010"

d1 = d1_raw[!is.na(numerise(d1_raw[, yearname])),][,c("country", yearname)]
colnames(d1)[2] = "lngdp"
d2 = d2_raw[!is.na(numerise(d2_raw[, yearname])),][,c("country", yearname)]
colnames(d2)[2] = "snt"
d3 = d3_raw[!is.na(numerise(d3_raw[, yearname])),][,c("country", yearname)]
colnames(d3)[2] = "lfx"

dtemp = merge(x = d1, y = d2, by = "country")
d = merge(x = dtemp, y = d3, by = "country")

d$lngdp = log(numerise(d$lngdp))

write.csv(d, "./Data/assembled.csv")

kable(head(d, 6L))
```

::: aside
<sub>FREE DATA FROM [UN](un.org), [WORLD BANK](https://worldbank.org), [WHO](https://who.org), [IMHE](http://www.healthdata.org/) VIA [GAPMINDER.ORG](https://gapminder.org), [CC-BY LICENSE](https://creativecommons.org/licenses/by/2.0/).</sub>
:::

# Univariate Statistics

## Mean and Standard Deviation {.smaller}

*Sample Mean* $\bar{x}$ is a measure of central tendency of a random variable $x$.

*Standard deviation* $s_x$ is a measure of dispersion in random variable $x$.

Note that $x_i$ is the i<sup>th</sup> observation of the random variable $x$. 

$$
\begin{align}
\bar{x} = {\frac{\sum _{i=1}^{n}(x_{i})}{n}}
&&
s_x = \sqrt{\frac{\sum_{i=1}^{n} \left(x_i - \bar{x}\right)^2}{n}}
\end{align}
$$

```{r echo=TRUE}

d_cor = data.frame(
  row.names = "Variable",
  Variable = c(
    "*ln(GDP)*",
    "*Sanitation*",
    "*Life Exp.*"
  ),
  Mean = c(
    mean(d$lngdp),
    mean(d$snt),
    mean(d$lfx)
  ),
  SD = c(
    sd(d$lngdp),
    sd(d$snt),
    sd(d$lfx)
  )
)

kable(
  d_cor,
  col.names = c(
    "Mean $\\bar{x}$",
    "Standard Deviation $s_x$"
  ),
  digits=5
)
```

# Scatter Plot

A *Scatter plot* is a type of Plot using Cartesian coordinate system to display values for two variables for a set of data. The data are displayed as a collection of points, each having one variable determining the *abscissa* and the other variable determining the *ordinate*. It helps us:

- take a short glance at effect of two variables.
- suggest kinds of correlations between variables.
- estimate the direction of correlation.

``` {r echo=TRUE}
sctrplot = function(
    d, x_map, y_map,
    x_lab=waiver(), y_lab=waiver(),
    title=waiver()
){
  plot1 = ggplot(d, mapping = aes(x = x_map, y = y_map))+
    geom_point(
      alpha=0.6
    )+
    theme_minimal()+
    theme(
      plot.background = element_rect(fill = "#f9f5d7", color = NA),
      panel.grid = element_line(colour = "#d5c4a1"),
      axis.line = element_line(colour = "#928374")
      )+
    labs(
      x=x_lab,
      y=y_lab,
      title=title
    )
  
  return(plot1)
}
```

## Sanitation vs. GDP per Capita

```{r}
sctrplot(d, d$lngdp, d$snt, "log of GDP per capita", "Sanitation Access %")
```

## Life Expectancy vs. GDP per Capita
```{r}
sctrplot(d, d$lngdp, d$lfx, "log of GDP per capita", "Life Expectancy")
```

## Life Expectation vs. Sanitation
```{r}
sctrplot(d, d$snt, d$lfx, "Sanitation Access %", "Life Expectancy")
```

# Bivariate Statistics

## Covariance and Correlation Matrices {.smaller}

*Covariance* $\operatorname{cov}(x, y)$ is a measure of the joint variability of two random variables $x$, $y$.

*Correlation* $r_{x,y}$ is any relationship, causal or spurious, between two random variables $x$, $y$. *Pearson's* $r$ correlation coefficient is considered here.

$$
\begin{align}
\operatorname {cov} (x,y)={\frac {\sum _{i=1}^{n}(x_{i}-\bar{x})(y_{i}-\bar{y})}{n}}
&&
r_{x,y}= \frac{\operatorname{cov}(x,y)}{s_x s_x}
\end{align}
$$

:::: {.columns}

::: {.column width="50%"}

```{r echo=TRUE}
cov_mat = cov(d[, 2:4])

kable(cov_mat, digits=5)
```

$$A_{i,j} = \operatorname{cov}(x_i, x_j)$$
:::

::: {.column width="50%"}

```{r echo=TRUE}
cor_mat = cor(d[, 2:4])

kable(cor_mat, digits=5)
```

$$A_{i,j} = r_{x_i, x_j}$$
:::

::::

## Other Correlation Coefficients

```{r echo=TRUE}
d_cor = data.frame(
  row.names = "Variable",
  Variable = c(
    "*Sanitation vs. ln(GDP)*",
    "*Life Exp. vs. ln(GDP)*",
    "*Life Exp. vs. Sanitation*"
  ),
  Pearson = c(
    cor(d$snt, d$lngdp, method="pearson"),
    cor(d$lfx, d$lngdp, method="pearson"),
    cor(d$lfx, d$snt, method="pearson")
  ),
  Spearman = c(
    cor(d$snt, d$lngdp, method="spearman"),
    cor(d$lfx, d$lngdp, method="spearman"),
    cor(d$lfx, d$snt, method="spearman")
  ),
  Kendall = c(
    cor(d$snt, d$lngdp, method="kendall"),
    cor(d$lfx, d$lngdp, method="kendall"),
    cor(d$lfx, d$snt, method="kendall")
  )
)

kable(
  d_cor,
  digit = 5,
  col.names = c(
    "*Pearson's* $r$",
    "*Spearman's* $r_s$",
    "*Kendall's* $\\tau$"
  )
)
```

# Linear Regression

*Simple Univariate Linear Regression*  is a method for estimating the relationship $y_i=f(x_i)$ of a *response* variable $y$ with a *predictor* variable $x$, as a line that closely fits the $y$ vs. $x$ *scatter plot*.

$$
y_i = \hat{a} + \hat{b} x_i + e_i.
$$

Where $\hat{a}$ is the *intercept*, $\hat{b}$ is the *slope*, and $e_i$ is the i<sup>th</sup> residual *error*. We aim to minimize $e_i$ for better fit.

## Ordinary Least Squares

*Ordinary Least squares method* reduces $e_i$ by minimizing *error sum of squares* $\sum{e_i^2}$.

```{r echo=TRUE}
olssmry = function(
    d, x_map, y_map,
    x_lab=waiver(), y_lab=waiver(),
    title=waiver()
){
  model = lm(formula=y_map~x_map)
  smry = summary(model, signif.stars=TRUE)
  
  smryvec = c(
    as.numeric(model$coefficients["(Intercept)"]),
    as.numeric(model$coefficients["x_map"]),
    smry$r.squared
  )
  
  return(smryvec)
}

olstab = t(data.frame(
  SvG = olssmry(d, d$lngdp, d$snt),
  LvG = olssmry(d, d$lngdp, d$lfx),
  LvS = olssmry(d, d$snt, d$lfx)
))

row.names(olstab) = c(
  "*Sanitation vs. ln(GDP)*",
  "*Life Exp. vs. ln(GDP)*",
  "*Life Exp. vs. Sanitation*"
)

kable(
  olstab,
  digit = 5,
  col.names=c(
  "$\\hat{a}$",
  "$\\hat{b}$",
  "$R^2$"
  )
)
```

::: aside
$R^2$: Coefficient of Determination
:::

## Least Absolute Deviation

*Least absolute Deviation method* reduces $e_i$ by minimizing the *sum of absolute deviations* $\sum{|e_i|}$.

```{r echo=TRUE}
ladsmry = function(
    d, x_map, y_map,
    x_lab=waiver(), y_lab=waiver(),
    title=waiver()
){
  model = rq(formula=y_map~x_map)
  smry = summary(model)
  
  smryvec = c(
    as.numeric(model$coefficients[1]),
    as.numeric(model$coefficients[2])
  )
  
  return(smryvec)
}

olstab = t(data.frame(
  SvG = ladsmry(d, d$lngdp, d$snt),
  LvG = ladsmry(d, d$lngdp, d$lfx),
  LvS = ladsmry(d, d$snt, d$lfx)
))

row.names(olstab) = c(
  "*Sanitation vs. ln(GDP)*",
  "*Life Exp. vs. ln(GDP)*",
  "*Life Exp. vs. Sanitation*"
)

kable(
  olstab,
  digit = 5,
  col.names=c(
  "$\\hat{a}$",
  "$\\hat{b}$"
  )
)
```

## Line fitting

Plotting the estimated *Linear Model* on the Scatter Plot.

```{r echo=TRUE}
linearplot = function(
    d, x_map, y_map,
    x_lab=waiver(), y_lab=waiver(),
    title=waiver()
){
  plot1 = ggplot(d, mapping = aes(x = x_map, y = y_map))+
    geom_point(
      alpha=0.6
    )+
    theme_minimal()+
    theme(
      plot.background = element_rect(fill = "#f9f5d7", color = NA),
      panel.grid = element_line(colour = "#d5c4a1"),
      axis.line = element_line(colour = "#928374")
      )+
    labs(
      x=x_lab,
      y=y_lab,
      title=title
    )+
    geom_smooth(
      method="lm",
      formula=y~x,
      se=FALSE,
      aes(color = "Ordinary Least Squares")
    )+
    geom_smooth(
      method="rq",
      formula=y~x,
      se=FALSE,
      aes(color = "Least Absolute Deviation")
    )+
    labs(
      color="Linear Model"
    )+
    scale_color_manual(
      values=c(
        "#cc241d80",
        "#45858880")
      )
  
  return(plot1)
}
```

## Sanitation vs. GDP per Capita

```{r}
linearplot(d, d$lngdp, d$snt, "log of GDP per capita", "Sanitation Access %")
```

## Life Expectancy vs. GDP per Capita
```{r}
linearplot(d, d$lngdp, d$lfx, "log of GDP per capita", "Life Expectancy")
```

## Life Expectation vs. Sanitation
```{r}
linearplot(d, d$snt, d$lfx, "Sanitation Access %", "Life Expectancy")
```

# Partial Correlation