---
title: Quarto Learning
author: Raj Pratap Singh, Shreyansh Mukhopadhyay, Aman Das
format:
  revealjs:
    theme: simple
    transition: slide
---

# intro
This is me learning to use quarto for the *first* time. I want to create a **presentation** using `ggplot2`
functions. Data to source from [gapminder](https://gapminder.org) and learnt the subject from <https://www.isical.ac.in>

## graph
```{r, echo=FALSE}
library(ggplot2)
```

```{r echo=TRUE}
ggplot(CO2, aes(x=conc,
                y = uptake))+
  geom_point()
```
---
```{r}
mean_uptake = mean(CO2$uptake)
```
The average uptake is `r mean_uptake`.

## Equation

$$
\Large
e^{i\pi} - 1 = 0
$$

# Tabbed

## Scatterplot
::: {.panel-tabset}

### Graph

```{r echo=FALSE}
ggplot(CO2, aes(x=conc,
                y = uptake))+
  geom_point()
```

### Code

```{r echo=TRUE, eval=FALSE}
ggplot(CO2, aes(x=conc,
                y = uptake))+
  geom_point()
```

### Theory

A `scatter plot` (also called a `scatterplot`, `scatter graph`, `scatter chart`, `scattergram`, or `scatter diagram`) is a type of **Plot** or **mathematical diagram** using *Cartesian coordinate system* to display values for typically two Variable for a set of data. If the points are coded (color/shape/size), one additional variable can be displayed.
The data are displayed as a collection of points, each having the value of one variable determining the position on the horizontal axis and the value of the other variable determining the position on the **vertical axis**.
:::