---
title: End-Semester Project
subtitle: Bivariate Analysis between GDP per capita, Sanitation and Life Expectancy across Nations in 2010
author:
  - Aman Das [BS2206]
  - Raj Pratap Singh [BS2219]
  - Shreyansh Mukhopadhyay [BS2147]
format:
  revealjs:
    width: 1200
    height: 750
    output-file: PPT_Reveal.html
    embed-resources: true
    theme: [default, custom.scss]
    highlight-style: gruvbox
    transition: slide
    toc: true
    toc-depth: 1
    hide-inactive-cursor: true
    code-fold: true
    menu: false
    scrollable: true
    history: true
    preview-links: true
    controls: true
  html:
    output-file: DOC_html.html
    embed-resources: true
    toc: true
    html-math-method: katex
    highlight-style: gruvbox-dark
    code-fold: true
    theme:
      light: sandstone
      dark: slate
  pptx:
    output-file: PPT_powerpoint.pptx
    reference-doc: ./Templates/template1.pptx
  pdf:
    documentclass: article
    margin-left: 40pt
    margin-right: 80pt
    echo: false
    output-file: DOC_pdf.pdf
    toc: true
    number-sections: true
    code-fold: true
    highlight-style: gruvbox
execute:
  cache: true
---

```{r include=FALSE}
library(knitr)
library(ggplot2)
library(quantreg)
library(ppcor)
library(psych)
library(moments)
library(latex2exp)
library(maps)

require(data.table)

mytheme=theme_minimal()+
theme(
  plot.background = element_rect(fill = "#f9f5d7", color = NA),
  panel.grid = element_line(colour = "#d5c4a1"),
  axis.line = element_line(colour = "#928374"),
  )
mycolor=scale_color_manual(
      values=c(
        "#cc241d",
        "#458588",
        "#689d6a",
        "#d65d0e"
        ),
      aesthetics=c(
        "fill",
        "colour"
      )
      )
```

# Introduction

## Overview {.smaller}

Our *Aim* is to determine which features more directly affect the *Life Expectation* of the citizens.

```{r message=FALSE, warning=FALSE, cache=TRUE}
#| fig-align: center

plotmap = function(d3, legendname, yearnamelabel, color1, color2){
  
  world=map_data("world")

  w3<-data.frame(world,data=sample(10,length(unique(world$group)),T)[world$group])
  
  d3[d3 == "United States"] = "USA"
  d3[d3 == "Congo, Dem. Rep."] = "Democratic Republic of the Congo"
  d3[d3 == "Congo, Rep."] = "Republic of Congo"
  d3[d3 == "Kyrgyz Republic"] = "Kyrgyzstan"
  d3[d3 == "Cote d'Ivoire"] = "Ivory Coast"
  d3[d3 == "United Kingdom"] = "UK"
  
  dt_d3 = as.data.table(d3)
  setkey(dt_d3, country)
  dt_w = as.data.table(world)
  setkey(dt_w, region)
  
  new_w = as.data.frame(dt_w[dt_d3])
  
  ggplot(new_w,aes(long,lat,group=group,fill=data))+
      geom_polygon(color="#282828", linewidth=rel(0.5))+
      scale_fill_gradient(low=color1,high=color2)+
      mytheme+
    theme(
      axis.text=element_blank(),
      axis.title = element_blank(),
      axis.line=element_blank(),
      plot.caption = element_text(hjust = 0, color="#504945"),
      legend.position = "bottom",
      legend.justification = "right",
      legend.title = element_text(size=rel(1.5), vjust=0.9)
    )+
    labs(fill=legendname)
}

numerise = function(x){
  x[grepl("k$", x)] <- as.numeric(sub("k$", "", x[grepl("k$", x)]))*10^3
  x <- as.numeric(x)
  return(x)
}

yearname = "X2010"

d1_raw = read.csv(file.path(".","Data","gdp.csv"), fileEncoding = 'UTF-8-BOM')
d2_raw = read.csv(file.path(".","Data","sanitation.csv"), fileEncoding = 'UTF-8-BOM')
d3_raw = read.csv(file.path(".","Data","life_expectancy.csv"), fileEncoding = 'UTF-8-BOM')

yearname = "X2010"

d1 = d1_raw[!is.na(numerise(d1_raw[, yearname])),][,c("country", yearname)]
colnames(d1)[2] = "lngdp"
d2 = d2_raw[!is.na(numerise(d2_raw[, yearname])),][,c("country", yearname)]
colnames(d2)[2] = "snt"
d3 = d3_raw[!is.na(numerise(d3_raw[, yearname])),][,c("country", yearname)]
colnames(d3)[2] = "lfx"
```

::: columns
::: {.column width="33%"}
```{r fig.width=5, fig.height=3, fig.align='left'}
d1$data = log(numerise(d1$lngdp))
plotmap(d1, "ln(GDP)", "2010", "#504945", "#cc241d")
```

```{r fig.width=5, fig.height=3, fig.align='left'}
d2$data = d2$snt
plotmap(d2, "Sanitation", "2010", "#504945", "#458588")
```
:::

::: {.column width="66%"}
$$
~
$$

```{r fig.width=10, fig.height=5, fig.align='left'}
d3$data = d3$lfx
plotmap(d3, "Life Exp.", "2010", "#504945", "#689d6a")
```
:::
:::

## Variables {.smaller}

This presentation used various methods of *Bivariate Analysis* to infer relationship between various features of Nations.

- **log of GDP per capita**: Logarithm (base $e$) of Gross Domestic Product (in \$) per citizen. Adjusted for Inflation. \[*lngdp*\]

  The Gross Domestic Product per capita of a country basically tells about the wealth of the citizens of the country.
- **Sanitation Access %**: Percentage of people using at least basic Sanitation facilities, not shared with other households. \[*snt*\]

  Sanitation of oneself is one of the basic tasks of a human being. We know poor sanitation is linked to transmission of diarrheal diseases such as cholera and dysentery, as well as typhoid, intestinal worm infections and polio which directly affect the health of an individual.

- **Life Expectancy**: The average number of years a newly born child would live, provided current mortality patterns hold. \[*lfx*\]

  Life expectancy is calculated based on the assumption that probability of death at a certain age stays constant in future. Hence we Life Expectation as a measurable proxy for Health.

## Data

```{r echo=TRUE}
script.dir <- getSrcDirectory(function(x) {x})
setwd(script.dir)

numerise = function(x){
  x[grepl("k$", x)] <- as.numeric(sub("k$", "", x[grepl("k$", x)]))*10^3
  x <- as.numeric(x)
  return(x)
}

d1_raw = read.csv(file.path(".","Data","gdp.csv"), fileEncoding = 'UTF-8-BOM')
d2_raw = read.csv(file.path(".","Data","sanitation.csv"), fileEncoding = 'UTF-8-BOM')
d3_raw = read.csv(file.path(".","Data","life_expectancy.csv"), fileEncoding = 'UTF-8-BOM')

yearname = "X2010"

d1 = d1_raw[!is.na(numerise(d1_raw[, yearname])),][,c("country", yearname)]
colnames(d1)[2] = "lngdp"
d2 = d2_raw[!is.na(numerise(d2_raw[, yearname])),][,c("country", yearname)]
colnames(d2)[2] = "snt"
d3 = d3_raw[!is.na(numerise(d3_raw[, yearname])),][,c("country", yearname)]
colnames(d3)[2] = "lfx"

dtemp = merge(x = d1, y = d2, by = "country")
d = merge(x = dtemp, y = d3, by = "country")

d$lngdp = log(numerise(d$lngdp))

write.csv(d, "./Data/assembled.csv")

kable(head(d, 6L))
```

::: aside
<sub>FREE DATA FROM [UN](un.org), [WORLD BANK](https://worldbank.org), [WHO](https://who.org), [IMHE](http://www.healthdata.org/) VIA [GAPMINDER.ORG](https://gapminder.org), [CC-BY LICENSE](https://creativecommons.org/licenses/by/2.0/).</sub>
:::

# Elementary Univariate Analysis

## Measures of Central Tendency {.smaller}

*Mean or Arithmetic Mean* $\bar{x}$, *Median* $\operatorname{median}(x)$ and *Mode* $\operatorname{mode}(x)$ are some measures of *central tendency* in the sample.

<details>

<summary>Formulae</summary>

$$
\begin{aligned}
x = \{x_1, x_2, \ldots,x_{n-1}, x_n\} &&
\operatorname{mean}(x)=\bar{x} = \frac{1}{n} \sum _{i=1}^{n}(x_{i})
\end{aligned}
$$$$
\begin{aligned}
\operatorname{median}(x)= \frac{x_{\lfloor\frac{n+1}{2}\rfloor}+x_{\lfloor\frac{n+2}{2}\rfloor}}{2} &&
\operatorname{mode}(x) = x_i \text{ s.t. } \operatorname{Pr}(x_i) = \operatorname{sup}(\operatorname{Pr}(x))  
\end{aligned}
$$

Note: $f_i$ is the frequency of the i<sup>th</sup> observation. $x_{(i)}$ is the i<sup>th</sup> largest observation.

</details>

```{r echo=TRUE}
getmode <- function(v) {
 uniqv <- unique(v)
 freq = max(tabulate(match(v, uniqv)))
 res = uniqv[which.max(tabulate(match(v, uniqv)))]
 if (freq == 1) res = NULL
 return(res)
}

d_central = data.frame(
  row.names = "Variable",
  Variable = c(
    "*ln(GDP)*",
    "*Sanitation*",
    "*Life Exp.*"
  ),
  Mean = c(
    mean(d$lngdp),
    mean(d$snt),
    mean(d$lfx)
  ),
  Median = c(
    median(d$lngdp),
    median(d$snt),
    median(d$lfx)
  ),
  Mode = c(
    getmode(d$lngdp),
    getmode(d$snt),
    getmode(d$lfx)
  )
)

kable(
  d_central,
  col.names = c(
    "$\\quad \\quad \\bar{x}$",
    "$\\operatorname{median}(x)$",
    "$\\operatorname{mode}(x)$"
  ),
  digits=5
)
```

$~$

-   Notice that mode of Sanitation is 100. Thus a large number of countries have universal access to basic sanitation infrastructure.

## Measures of Dispersion {.smaller}

*Range* $\operatorname{range}(x)$, *Semi-Interquartile Range* $\operatorname{SIR}(x)$, *Mean Deviation about x'* $\operatorname{MD}_{(x')}(x)$, *Variance* $s_x^2$, *Standard Deviation* $s_x$ are some measures of *dispersion* in the sample.

<details>

<summary>Formulae</summary>

$$
\begin{aligned}
\operatorname{range}(x)=|x_{(n)} - x_{(1)}| &&
\ Q_1 = \operatorname{median}(x_{(1)}, \ldots ,x_{(\lfloor \frac{n+1}{2} \rfloor)}) &&
\end{aligned}
$$ $$
\begin{aligned}
\ Q_3 = \operatorname{median}(x_{(\lfloor \frac{n+2}{2} \rfloor)}, \ldots , x_{(n)}) &&
\operatorname{MD}_{(x')}(x) = \operatorname{mean}(|x_i-x'|)
\end{aligned}
$$ $$
\begin{aligned}
\operatorname{SIR}(x)=\frac{|Q_1-Q_3|}{2} &&
s_x = \sqrt{\operatorname{mean}([x_i - \bar{x}]^2)} &&
s^2_x= (s_x)^2
\end{aligned}
$$

</details>

```{r echo=TRUE}
getmd = function(x, center = mean(x)){
  md = mean(
    abs(
      x - rep(center, length(x))
      )
    )
  return(md)
}
d_disp = data.frame(
  row.names = "Variable",
  Variable = c(
    "*ln(GDP)*",
    "*Sanitation*",
    "*Life Exp.*"
  ),
  Range = c(
    max(d$lngdp) - min(d$lngdp),
    max(d$snt) - min(d$snt),
    max(d$lfx) - min(d$lfx)
  ),
  SIR = c(
    IQR(d$lngdp)/2,
    IQR(d$snt)/2,
    IQR(d$lfx)/2
  ),
  MD = c(
    getmd(d$lngdp),
    getmd(d$snt),
    getmd(d$lfx)
  ),
  variance = c(
    (sd(d$lngdp))^2,
    (sd(d$snt))^2,
    (sd(d$lfx))^2
  ),
  SD = c(
    sd(d$lngdp),
    sd(d$snt),
    sd(d$lfx)
  )
)

kable(
  d_disp,
  col.names = c(
    "$\\operatorname{range}(x)$",
    "$\\operatorname{SIR}(x)$",
    "$\\operatorname{MD}_{(\\bar{x})}(x)$",
    "$\\quad \\quad \\quad \\quad s_x^2$",
    "$\\quad \\quad \\quad \\quad s_x$"
  ),
  digits=5
)
```

$~$

-   We can compare $s_x$ to $\bar{x}$ and observe that there is a high variation in Sanitation amongst countries.
-   GDP per capita varies drastically across the $e$<sup>`r max(d$lngdp) - min(d$lngdp)`</sup> range.

## Measures of Shape {.smaller}

Coefficients of *Skewness* $g_1$ and *Kurtosis* $g_2$ describe the symmetry and extremity of tails of the sample distribution.

<details>

<summary>Formulae</summary>

$$
\begin{aligned}
m_k = \operatorname{mean}([x-\bar{x}]^k) &&
g_1 = \frac{m_3}{{m_2}^{\frac{3}{2}}} &&
g_2 = \frac{m_4}{{m_2}^2}
\end{aligned}
$$

</details>

```{r echo=TRUE}
d_shape = data.frame(
  row.names = "Variable",
  Variable = c(
    "*ln(GDP)*",
    "*Sanitation*",
    "*Life Exp.*"
  ),
  Skewness = c(
    skewness(d$lngdp),
    skewness(d$snt),
    skewness(d$lfx)
  ),
  Kurtosis = c(
    kurtosis(d$lngdp),
    kurtosis(d$snt),
    kurtosis(d$lfx)
  )
)

kable(
  d_shape,
  col.names = c(
    "Skewness $g_1$",
    "Kurtosis $g_2$"
  ),
  digits=5
)
```

$$
~
$$

- ln(GDP) is nearly symmetrical, while Sanitation and Life Exp. are highly left-skewed.

  This indicates majority of countries have good sanitation system and citizen health.
  
- ln(GDP) and Sanitation are platykurtic, while Life Exp. is leptokurtic.


## Density Plot {.smaller}

```{r echo=TRUE}
#| fig-align: center
labelfunction = function(val1){
  return(list(c(
    "log of GDP per capita",
    "Sanitation Access %",
    "Life Expectancy"
  )))
}
ggplot(stack(d[2:4]), mapping = aes(x = values))+
geom_density(aes(color=ind), linewidth=rel(1.5))+
labs(
  x=NULL,
  y=NULL
  )+
mytheme+
mycolor+
facet_wrap(~ind, scales="free", labeller = labelfunction, ncol=1)+
  theme(legend.position="none",
        strip.text.x = element_text(size = rel(1.5))
  )
```

- The density of the log of GDP per capita is symmetric as inferred previously. The other two density plots appear left skewed as supported by the negative skewness.

## Box Plot {.smaller scrollable="false"}

*Box plots* help us detect potential outliers. They also help us in estimating location and skewness of the distribution.

```{r echo=TRUE}
#| fig-align: center
labelfunction = function(val1){
  return(list(c(
    "log of GDP per capita",
    "Sanitation Access %",
    "Life Expectancy"
  )))
}
ggplot(stack(d[2:4]), mapping = aes(y = values))+
geom_boxplot(aes(fill=ind), alpha=0.6)+
labs(
  x=NULL,
  y=NULL
  )+
mytheme+
mycolor+
facet_wrap(~ind, scales="free", labeller = labelfunction)+
  theme(axis.text.x=element_blank(),
        legend.position="none",
        strip.text.x = element_text(size = rel(1.5)),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()
  )
```

-   We observe one potential outlier in the Life Expectancy dataset. It is Haiti at `r subset(d, country=="Haiti")[4]` years. It is due to a cholera outbreak and an earthquake increasing the mortality rate in the nation in 2010.

# Scatter Plot

A *Scatter plot* helps us estimate the type of relationship between variables.

```{r echo=TRUE}
sctrplot = function(
    d, x_map, y_map,
    x_lab=waiver(), y_lab=waiver(),
    title=waiver()
){
  plot1 = ggplot(d, mapping = aes(x = x_map, y = y_map))+
    geom_point(
      alpha=0.6
    )+
    mytheme+
    labs(
      x=x_lab,
      y=y_lab,
      title=title
    )
  
  return(plot1)
}
```

## Sanitation vs. log of GDP {.smaller}

```{r}
sctrplot(d, d$lngdp, d$snt, "log of GDP per capita", "Sanitation Access %")
```

- We observe from the scatter plot that the correlation between ln(GDP) and sanitation access appears to be mostly Linear for countries with low ln(GDP).
- Also, the sanitation access is very close to 100% for countries with high ln(GDP).

## Life Expectancy vs. log of GDP {.smaller}

```{r}
sctrplot(d, d$lngdp, d$lfx, "log of GDP per capita", "Life Expectancy")
```

- We can see that except for some countries with very low life expectancies, the correlation between the variables is appearing to be linear.

## Life Expectation vs. Sanitation {.smaller}

```{r}
sctrplot(d, d$snt, d$lfx, "Sanitation Access %", "Life Expectancy")
```

- We can see that the correlation appears to be linear between life expectation and sanitation. We also observe that there is clustering around the top right side of the plot, which is supported by the box plots of both the distributions.

---

### Inferences

We observe a fairly strong positive Linear Correlation between the three features in the Scatter Plot.

We subsequently compute the Correlation Coefficients to quantify the Linear Correlation.

# Bivariate Statistics

## Correlation Coefficients {.smaller}

*Correlation* is any relationship, causal or spurious, between two random variables $x$, $y$.

*Pearson's* $r$, *Spearman's* $r_s$, and *Kendall's* $\tau$ are some correlation coefficients. These estimate the linear correlation between two variables.

```{r echo=TRUE}
d_cor = data.frame(
  row.names = "Variable",
  Variable = c(
    "*Sanitation vs. ln(GDP)*",
    "*Life Exp. vs. ln(GDP)*",
    "*Life Exp. vs. Sanitation*"
  ),
  Pearson = c(
    cor(d$snt, d$lngdp, method="pearson"),
    cor(d$lfx, d$lngdp, method="pearson"),
    cor(d$lfx, d$snt, method="pearson")
  ),
  Spearman = c(
    cor(d$snt, d$lngdp, method="spearman"),
    cor(d$lfx, d$lngdp, method="spearman"),
    cor(d$lfx, d$snt, method="spearman")
  ),
  Kendall = c(
    cor(d$snt, d$lngdp, method="kendall"),
    cor(d$lfx, d$lngdp, method="kendall"),
    cor(d$lfx, d$snt, method="kendall")
  )
)

avg_cor = round(mean(d_cor[, 1]), digits=2)

kable(
  d_cor,
  digit = 5,
  col.names = c(
    "*Pearson's* $r$",
    "*Spearman's* $r_s$",
    "*Kendall's* $\\tau$"
  )
)
```

## Covariance and Correlation Matrices {.smaller}

*Covariance* $\operatorname{cov}(x, y)$ is a measure of the joint variability of two random variables $x$, $y$.

<details>

<summary>Formulae</summary>

$$
\begin{aligned}
\operatorname {cov} (x,y)={\frac {\sum _{i=1}^{n}(x_{i}-\bar{x})(y_{i}-\bar{y})}{n}}
&&
r_{x,y}= \frac{\operatorname{cov}(x,y)}{s_x s_x}
\end{aligned}
$$

</details>

::: columns
::: {.column width="48%"}
```{r echo=TRUE}
cov_mat = cov(d[, 2:4])

kable(cov_mat, digits=5)
```

$$A_{i,j} = \operatorname{cov}(x_i, x_j)$$
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
```{r echo=TRUE}
cor_mat = cor(d[, 2:4])

kable(cor_mat, digits=5)
```

$$A_{i,j} = r_{x_i, x_j}$$
:::
:::

---

### Inferences

We observe fairly good Linear Correlation of around `r avg_cor` between all three variables. Thus a slight increase in ln(GDP), Sanitation or Life expectancy tends to accompany rise in the other two features increasing too.

Let us try to create a Linear Model to best estimate the relationship between the three variables.

# Regression

## Simple Linear Regression

*Simple Univariate Linear Regression* is a method for estimating the relationship $y_i=f(x_i)$ of a *response* variable $y$ with a *predictor* variable $x$, as a line that closely fits the $y$ vs. $x$ *scatter plot*.

$$
y_i = \hat{a} + \hat{b} x_i + e_i
$$

Where $\hat{a}$ is the *intercept*, $\hat{b}$ is the *slope*, and $e_i$ is the i<sup>th</sup> residual *error*. We aim to minimize $e_i$ for better fit.

## Ordinary Least Squares {.smaller}

*Ordinary Least squares method* reduces $e_i$ by minimizing *error sum of squares* $\sum{{e_i}^2}$.

*Coefficient of Determination* $R^2$ is the proportion of the variation in $y$ predictable by the model.

<details>

<summary>Formulae</summary>

$$
\begin{aligned}
\hat{b} = r\frac{s_y}{s_x} &&
\hat{a} = \bar{y} - \hat{b}\bar{x} &&
R^2 = 1 - \frac{\sum{{e_i}^2}}{\sum{(y-\bar{y}})^2}
\end{aligned}
$$

</details>

```{r echo=TRUE}
olssmry = function(
    d, x_map, y_map,
    x_lab=waiver(), y_lab=waiver(),
    title=waiver()
){
  model = lm(formula=y_map~x_map)
  smry = summary(model, signif.stars=TRUE)
  
  smryvec = c(
    as.numeric(model$coefficients["(Intercept)"]),
    as.numeric(model$coefficients["x_map"]),
    smry$r.squared
  )
  
  return(smryvec)
}

olstab = t(data.frame(
  SvG = olssmry(d, d$lngdp, d$snt),
  LvG = olssmry(d, d$lngdp, d$lfx),
  LvS = olssmry(d, d$snt, d$lfx)
))

row.names(olstab) = c(
  "*Sanitation vs. ln(GDP)*",
  "*Life Exp. vs. ln(GDP)*",
  "*Life Exp. vs. Sanitation*"
)

avg_r2 = round(mean(olstab[, 3]), digits = 1)

kable(
  olstab,
  digit = 5,
  col.names=c(
  "$\\hat{a}$",
  "$\\hat{b}$",
  "$R^2$"
  )
)
```

- The $R^2$ values of all three models are all near `r avg_r2` . Thus the models explain the variation in the response $y_i$ fairly well.

## Least Absolute Deviation

*Least absolute Deviation method* reduces $e_i$ by minimizing the *sum of absolute deviations* $\sum{|e_i|}$.

```{r echo=TRUE}
ladsmry = function(
    d, x_map, y_map,
    x_lab=waiver(), y_lab=waiver(),
    title=waiver()
){
  model = rq(formula=y_map~x_map)
  smry = summary(model)
  
  smryvec = c(
    as.numeric(model$coefficients[1]),
    as.numeric(model$coefficients[2])
  )
  
  return(smryvec)
}

olstab = t(data.frame(
  SvG = ladsmry(d, d$lngdp, d$snt),
  LvG = ladsmry(d, d$lngdp, d$lfx),
  LvS = ladsmry(d, d$snt, d$lfx)
))

row.names(olstab) = c(
  "*Sanitation vs. ln(GDP)*",
  "*Life Exp. vs. ln(GDP)*",
  "*Life Exp. vs. Sanitation*"
)

kable(
  olstab,
  digit = 5,
  col.names=c(
  "$\\hat{a}$",
  "$\\hat{b}$"
  )
)
```

# Line Fitting

Plotting the estimated *Linear Models* on the Scatter Plots.

```{r echo=TRUE}

linearplot = function(
    d, x_map, y_map,
    x_lab=waiver(), y_lab=waiver(),
    title=waiver()
){
  olsvec = round(olssmry(d, x_map, y_map), digit=5)
  ladvec = round(ladsmry(d, x_map, y_map), digit=5)
  capstr = TeX(paste("$y_i =", olsvec[1], "+", olsvec[2], "~x_i + e_i$", "\t\t",
                     "$y_i' =", ladvec[1], "+", ladvec[2], "~x_i + e_i'$"))
  
  plot1 = ggplot(d, mapping = aes(x = x_map, y = y_map))+
    geom_point(
      alpha=0.6
    )+
    mytheme+
    labs(
      x=x_lab,
      y=y_lab,
      title=title,
      caption=capstr,
      parse=TRUE
    )+
    geom_smooth(
      method="lm",
      formula=y~x,
      se=FALSE,
      aes(color = "Ordinary Least Squares")
    )+
    geom_smooth(
      method="rq",
      formula=y~x,
      se=FALSE,
      aes(color = "Least Absolute Deviation")
    )+
    labs(
      color="Linear Model"
    )+
    scale_color_manual(
      values=c(
        "#cc241d80",
        "#45858880",
        "#689d6a80",
        "#d65d0e80"
        )
      )+
    theme(
      plot.caption = element_text(hjust=0.5, color="#504945")
    )
  
  return(plot1)
}
```

## Sanitation vs. log of GDP
```{r warning=FALSE}
#| fig-align: center
linearplot(d, d$lngdp, d$snt, "log of GDP per capita", "Sanitation Access %")+
  ylim(0, 100)
```

## Life Expectancy vs. log of GDP

```{r warning=FALSE}
#| fig-align: center
linearplot(d, d$lngdp, d$lfx, "log of GDP per capita", "Life Expectancy")
```

## Life Expectancy vs. Sanitation

```{r warning=FALSE}
#| fig-align: center
linearplot(d, d$snt, d$lfx, "Sanitation Access %", "Life Expectancy")
```

---

### Inferences

Both OLS and LAD models fit the scatter plots very well. The OLS model is affected in the Life Expectancy plots due to the existence of Haiti as an outlier.

Our aim is to deduce which features more directly affect the Life Expectancy. Thus we should remove the effect of the other features when computing correlation.

# Partial Correlation {.smaller}

*Partial Correlation* is the relationship between two variables $x$, $y$ of interest, after removing effect of some other related variable $z$.

<details>

<summary>Formulae</summary>

$$
\begin{aligned}
&x_i = \hat{a_x} + \hat{b_x} z_i + e_{x,i} &&
y_i = \hat{a_y} + \hat{b_y} z_i + e_{y,i}
\\
&\Rightarrow r_{x,y;z} = r_{e_{x}, e_{y}}
\end{aligned}
$$

</details>

::: columns
::: {.column width="40%"}
```{r echo=TRUE}
partcor = pcor(d[, 2:4])$estimate

pcortab = data.frame(
  row.names = "Variable",
  Variable = c(
    "*Sanitation vs. ln(GDP)*",
    "*Life Exp.$\\quad$ vs. ln(GDP)*",
    "*Life Exp. vs. Sanitation*"
  ),
  PCor = c(
    partcor[2, 1],
    partcor[3, 1],
    partcor[3, 2]
  )
)

kable(pcortab,
      col.names = c(
        "Partial Correlation"
      ))
```
:::

::: {.column width="5%"}
:::

::: {.column width="55%"}
-   We observe that the partial correlation between Life Exp. and Sanitation is higher than that of Life. Exp and ln(GDP).

Thus Life Expection is more directly improved by better Sanitaion access, than increase in wealth.

-   We also note that there is fairly high partial correlation between Sanitation and ln(GDP).

This indicates that wealthier countries tend to improve their sanitation infrastructure.
:::
:::

# Considerations {.smaller}

## Time

```{r echo=TRUE, fig.width=20}

library(knitr)
library(ggplot2)
library(quantreg)
library(ppcor)
library(psych)
library(moments)
library(latex2exp)
library(maps)
library(dplyr)

numerise = function(x){
  x[grepl("k$", x)] <- as.numeric(sub("k$", "", x[grepl("k$", x)]))*10^3
  x <- as.numeric(x)
  return(x)
}

d1_raw = read.csv(file.path(".","Data","gdp.csv"), fileEncoding = 'UTF-8-BOM')
d2_raw = read.csv(file.path(".","Data","sanitation.csv"), fileEncoding = 'UTF-8-BOM')
d3_raw = read.csv(file.path(".","Data","life_expectancy.csv"), fileEncoding = 'UTF-8-BOM')

years = 2000:2019
yearnames = paste0("X", years)

makedata = function(yearname){
  d1 = d1_raw[!is.na(numerise(d1_raw[, yearname])),][,c("country", yearname)]
  colnames(d1)[2] = "lngdp"
  d2 = d2_raw[!is.na(numerise(d2_raw[, yearname])),][,c("country", yearname)]
  colnames(d2)[2] = "snt"
  d3 = d3_raw[!is.na(numerise(d3_raw[, yearname])),][,c("country", yearname)]
  colnames(d3)[2] = "lfx"
  
  dtemp = merge(x = d1, y = d2, by = "country")
  d = merge(x = dtemp, y = d3, by = "country")
  
  d$lngdp = log(numerise(d$lngdp))
  
  return(d)
}
```

<details>

<summary>How the years' life expectancies and an year's GDP correlate with.</summary>

```{r echo=FALSE, cache=TRUE}
par(mfcol=c(1,2), bg="#f9f5d7")
for (yearname0 in yearnames) {
  r2vector = c()
  d0 = makedata(yearname0)
  for (yearname1 in yearnames) {
  d1 = makedata(yearname1)
    countrynames = intersect(d0$country, d1$country)
    d1 = d1 %>% filter(country %in% countrynames)
    d0 = d0 %>% filter(country %in% countrynames)
    r2vector = append(r2vector, cor(d1$lfx, d0$lngdp))
  }
  plot(years, r2vector, xlab=paste(yearname0, "ln(GDP)"), ylab="Corr. of Life Exp.")
}
```

</details>

<details>

<summary>How the years' life expectancies and an year's Sanitation correlate with.</summary>

```{r echo=FALSE, cache=TRUE}
par(mfcol=c(1,2), bg="#f9f5d7")
for (yearname0 in yearnames) {
  r2vector = c()
  d0 = makedata(yearname0)
  for (yearname1 in yearnames) {
  d1 = makedata(yearname1)
    countrynames = intersect(d0$country, d1$country)
    d1 = d1 %>% filter(country %in% countrynames)
    d0 = d0 %>% filter(country %in% countrynames)
    r2vector = append(r2vector, cor(d1$lfx, d0$snt))
  }
  plot(years, r2vector, xlab=paste(yearname0, "Sanitation"), ylab="Corr. of Life Exp.")
}
```

</details>

<details>

<summary>How an year's life expectancies correlate with the years' GDP.</summary>

```{r echo=FALSE, cache=TRUE}
par(mfcol=c(1,2), bg="#f9f5d7")
for (yearname0 in yearnames) {
  r2vector = c()
  d0 = makedata(yearname0)
  for (yearname1 in yearnames) {
  d1 = makedata(yearname1)
    countrynames = intersect(d0$country, d1$country)
    d1 = d1 %>% filter(country %in% countrynames)
    d0 = d0 %>% filter(country %in% countrynames)
    r2vector = append(r2vector, cor(d0$lfx, d1$lngdp))
  }
  plot(years, r2vector, xlab=paste(yearname0, "Life Exp."), ylab="Corr. of ln(GDP)")
}
```

</details>

<details>

<summary>How an year's life expectancies correlate with the years' Sanitation.</summary>

```{r echo=FALSE, cache=TRUE}
par(mfcol=c(1,2), bg="#f9f5d7")
for (yearname0 in yearnames) {
  r2vector = c()
  d0 = makedata(yearname0)
  for (yearname1 in yearnames) {
  d1 = makedata(yearname1)
    countrynames = intersect(d0$country, d1$country)
    d1 = d1 %>% filter(country %in% countrynames)
    d0 = d0 %>% filter(country %in% countrynames)
    r2vector = append(r2vector, cor(d0$lfx, d1$lngdp))
  }
  
  plot(years, r2vector, xlab=paste(yearname0, "Life Exp."), ylab="Corr. of Sanitation")
}
```

</details>

Thus our analysis is fairly robust with respect to variation in Time. Thus we only perform the analysis on the year 2010.

## Raw GDP per capita

<details> <summary> Sanitation vs GDP per capita</summary>

```{r}
sctrplot(d, exp(d$lngdp), d$snt, "GDP per capita", "Sanitation Access %")
```
</details>

<details> <summary> Life Expectancy vs GDP per capita </summary>
```{r}
sctrplot(d, exp(d$lngdp), d$lfx, "GDP per capita", "Sanitation Access %")
```
</details>


Scatter Plots of Sanitation or Life Exp. with respect to raw GDP are not linearly correlated. As this was beyond our scope of knowledge, we instead considered the log (base $e$) of GDP.

Thus in reality, rise in GDP provides diminishing returns in citizen health the wealthier a country is.

# Conclusion {.scrollable}

Thus we have observed that Life expectancy increases with rise in log of GDP per capita and Sanitation facility access. Life Expectancy is also more directly affected by increase in basic Sanitation access than with rise in log of GDP per capita.

## Suggestions

Governments should consider that:

- Focusing on better sanitation services leads to visibly better lifespan of the citizens.

- Improvements in GDP per capita make practical improvements in sanitation and life expectancy only for poorer countries, due diminishing returns in the log scale.

- Improvements to life expectancies once made, are fairly stable.

## Notable Countries {.smaller}
Some countries had alarmingly low Life Expectancies:

- The countries like the Central African Republic, Zambia, Zimbabwe have very low life expectancies due to endemic poverty and weak governance, contributing to a dire health situation.

- Haiti had a high child mortality rate in 2010 due to natural disasters and cholera outbreaks. This caused low Life expectancy for that year.

```{r fig.height=3}
par(bg="#f9f5d7", fin=c(6, 3))
plot(2000:2019, t(subset(d3_raw,country == "Haiti"))[202:221], xlab = "Years", ylab = "Life Expectancy")
```

- Lesotho is experiencing an increasing double burden of diseases characterized by an increase in the burden of non-communicable diseases as well as a high burden of communicable diseases.

# Credits

-   Dr. Kiranmoy Das
-   [Gapminder](https://gapminder.org)
-   [Wikipedia](https://wikipedia.org)
-   [Statology](https://statology.org)
-   [Quarto](https://quarto.org)
-   [RStudio](https://posit.co/products/open-source/rstudio/)

# Thank You {visibility="uncounted"}